An efficient feedback calibration algorithm for direct imaging radio telescopes

### 2016-02-17 cal_demo/ms.tex
Further tweaking cal paper before sending to coauthors

% Enter the current year, for the copyright statements etc.
\pubyear{2015}

@@

Next we move toward the feedback calibration outlined in \citealt{mor11}. The calibration loop described there assumed a simple sky of a single point source, but we aim for a more generalized solution for arbitrarily complex sky models. As such the quantity we wish to inspect is a correlation of the uncalibrated antenna signals input to the correlator with the holographic electric field image output from the correlator. We define the quantity,
\begin{equation}\label{eq:Cna}
\Cna \equiv \left<\Er{a}(t) E'^*(\spix,t)\right>_t,
\end{equation}
where the superscript $n$ again represents the quantity formed in the $n^\mathrm{th}$ calibration loop, and \spix\, is the pixel center nearest a bright calibrator of interest. The following will hold for any chosen pixel, \spix, though it is advantageous to choose a pixel which contains a bright source to achieve a high signal to noise.

@@

Figure~\ref{fig:data_images} shows the improvement in the images due to our calibration. The left panel shows the uncalibrated image integrated over 51.2 ms, 29 kHz. Cyg~A is prominent near the center of the image, and Cas~A is also clearly visible in the upper right. The middle panel shows the image produced after calibration with identical integration time and bandwidth. The sidelobes throughout the image are significantly suppressed, and the galactic plane is much more evident. We also note that the feature just to the right of Cyg~A is dimmer in the calibrated image, better matching the expected flux from the global sky model (GSM, \citealt{deo08}). We show the GSM modulated by two factors of the LWA primary beam in the right panel of figure~\ref{fig:data_images} for reference.

\begin{figure*}
\begin{center}
\includegraphics[width=0.33\linewidth]{figures/cal_paper_data_image_before.pdf}
\includegraphics[width=0.33\linewidth]{figures/cal_paper_data_image_after.pdf}
\includegraphics[width=0.33\linewidth]{figures/cal_paper_gsm_beam_weighted.pdf}
\caption{Images produced before (left) and after (middle) calibrating LWA data. These images were produces with 29 kHz bandwidth and 51.2 ms integration. The calibrated image shows significant reduction in sidelobe rumble throughout, while retaining the prominent Cyg~A and Cas~A sources. The galactic plane is also substantially more evident after calibration. For reference, the GSM is shown in the same coordinates and weighted by two factors of the LWA primary beam in the right panel.
}
\label{fig:data_images}
\end{center}
\end{figure*}

@@

\section{Noise trends}\label{sec:noise}
Here we study the properties of our calibration solutions with respect to noise in the system, and compare to solutions found from visibilities. For this comparison we run a suite of simulations varying the receiver noise and integration times, while forming $\Cna$ correlations for EPICal gain solutions, and simultaneously forming visibilities from the same E-field streams to find visibility-based gain solutions.

We use a simulated sky consisting of a 5 Jy calibrator source, and 10 other random sources with apparent flux densities 0.2 to 0.5 Jy (total sky power $\approx$~8 Jy). We generated these sources randomly, but kept them fixed for each run. As in section~\ref{sec:sim}, we use the MWA core layout (51 antennas), and a 4.4~m square tophat antenna voltage pattern. We simulate 64 40~kHz frequency channels. Because our calibration loop is frequency independent, each channel can be treated as a separate trial of the simulation, and is used to better estimate the statistics.

@@

We initialize our gain estimates with the true values (unity), and allow the estimate to be corrupted by the noise through one iteration of the calibration loop. Ideally we would use a damping factor of zero to allow for the noise to fully corrupt the gains. However, we found that a zero damping factor resulted in non-convergent results, so we adopted a factor, $\damp=0.05$. For a fair comparison, we use the same factor for the visibility-based solutions. We then observe the error in the gain estimates introduced by the noise.
\begin{equation}\label{eq:gain_error}
\sigma_g = \left[\frac{1}{N_f \Nant} \sum_f \sum_a \frac{\left|g'_a(f)-g^T_a(f)\right|^2}{\left|g^T_a(f)\right|^2}\right]^{1/2}
\end{equation}

@@

\textbf{Improving sky model.} In this work we used a perfect sky model (for simulation), or a very simple sky model (for LWA data). In principle the images produced by the EPIC correlator can be used to improve the sky model used in calibration -- similar to a major loop in self-calibration. This can be especially useful for compact, widefield arrays which require a model of both compact and diffuse sources over a large patch of sky. In addition to requiring the sky model, the calibration also requires this model to be attenuated by the primary beam, which can be difficult to measure at the precision necessary \citep[e.g.][]{neb15,vir14,thy15b}. However, the direct imaging correlator provides exactly the model necessary in real time, and can be iterated over to improve these images and the gain solutions.

\textbf{Dynamic parameters.} In our controlled experiments we fine tuned a number of parameters based on our testing (e.g. damping factor, integration time, frequency averaging). A deployed system will require robust determination of these parameters to operate continuously. The specifics will be heavily dependent on the stability of the instrument, the frequency of observation, and the sky. For example, a stable instrument may be able to use short integrations to determine a rough estimate of the gains before switching to much longer integration (on order seconds to minutes) to highly increase signal to noise. At low frequencies or high sky resolution, the dynamics of the ionosphere are important, and will likely drive the limit of time integration allowed.

\textbf{Direction dependent gains.} Even below the timescale of a dynamic ionosphere, a widefield instrument may require direction dependent gains to account for a varying thickness ionospheric sheet. The holographic mapping framework used by the MOFF algorithm, and in turn the EPIC and EPICal software can naturally account for direction dependence through and updated model of the antenna voltage pattern. This can be achieved by first forming complex gain estimates toward multiple directions on the sky by selecting several pixels and independently following the EPICal algorithm for each. These solutions can then be used to fit a beam model on the sky (analogous to the gain model fitting mentioned above), and updating the antenna voltage pattern. This updated pattern would then feed into the gridding step of the correlator, allowing the imager to convolve the signals with the effective beam pattern in the ground plane. Because this method requires multiple pixel correlations, the scaling now becomes $\mathcal{O}(\Nant N_{\mathrm{pix}})$, which will typically be much lower than the $\mathcal{O}(\Ng \log_2 \Ng)$ of the correlator itself.

>>>>

% Enter the current year, for the copyright statements etc.
\pubyear{2016}

@@

Next we move toward the feedback calibration outlined in \citealt{mor11}. There it was suggested to form a correlation of the uncalibrated antenna measurements with an image pixel from the output of the correlator. This is a statistically stationary quantity, which can be related to the sum of visibilities involving the antenna used in the correlation -- exactly the sum needed to calibrate a simple sky of a single point source. However, we aim for a more generalized solution for arbitrarily complex sky models. We therefore study the full expression for the antenna-pixel correlation,
\begin{equation}\label{eq:Cna}
\Cna \equiv \left<\Er{a}(t) E'^*(\spix,t)\right>_t,
\end{equation}
where the superscript $n$ again represents the quantity formed in the $n^\mathrm{th}$ calibration loop, and \spix\, is the pixel center nearest a bright calibrator of interest. The following will hold for any chosen pixel, \spix, though it is advantageous to choose a pixel which contains a bright source to achieve a high signal to noise.

@@

Figure~\ref{fig:data_images} shows the improvement in the images due to our calibration. The left panel shows the uncalibrated image integrated over 51.2 ms, 29 kHz. Cyg~A is prominent near the center of the image, and Cas~A is also clearly visible in the upper right. The middle panel shows the image produced after calibration with identical integration time and bandwidth. The sidelobes throughout the image are significantly suppressed, and the galactic plane is much more evident, despite only modeling Cyg~A and Cas~A. We also note that the feature just to the right of Cyg~A is dimmer in the calibrated image, better matching the expected flux from the global sky model (GSM, \citealt{deo08}). We show the GSM modulated by two factors of the LWA primary beam in the right panel of figure~\ref{fig:data_images} for reference.

\begin{figure*}
\begin{center}
\includegraphics[width=0.33\linewidth]{figures/cal_paper_data_image_before.pdf}
\includegraphics[width=0.33\linewidth]{figures/cal_paper_data_image_after.pdf}
\includegraphics[width=0.33\linewidth]{figures/cal_paper_gsm_beam_weighted.pdf}
\caption{Images produced before (left) and after (middle) calibrating LWA data. These images were produces with 29 kHz bandwidth and 51.2 ms integration. The calibrated image shows significant reduction in sidelobe rumble throughout, while retaining the prominent Cyg~A and Cas~A sources. The galactic plane is also substantially more evident after calibration. For reference, the GSM is shown in the same coordinates and weighted by two factors of the LWA primary beam in the right panel. While much of the GSM is visible in the calibrated image, we note that the model used to calibrate was only the two bright sources, Cyg~A and Cas~A.
}
\label{fig:data_images}
\end{center}
\end{figure*}

@@

\section{Noise trends}\label{sec:noise}
Here we study the properties of our calibration solutions with respect to noise in the system, and compare to solutions found from visibilities. For this comparison we run a suite of simulations varying the receiver noise and integration times, while forming antenna-pixel correlations, $\Cna$, for EPICal gain solutions, and simultaneously forming visibilities from the same E-field streams to find visibility-based gain solutions.

We use a simulated sky consisting of a 5 Jy calibrator source, and 10 other random sources with apparent flux densities 0.2 to 0.5 Jy (total sky power $\approx$~8 Jy). We generated these sources randomly, but kept them fixed for each run. As in section~\ref{sec:sim}, we use the MWA core layout (51 antennas), and a 4.4~m square tophat antenna voltage pattern. We simulate 64 40~kHz frequency channels. Because our calibration loop treats frequency channels independently, each channel can be treated as a separate trial of the simulation, and is used to better estimate the statistics.

@@

We initialize our gain estimates with the true values (${g^{(0)}_a=g^T_a=1}$), and allow the estimate to be corrupted by the noise through one iteration of the calibration loop. Ideally we would use a damping factor of zero to allow for the noise to fully corrupt the gains. However, we found that a zero damping factor resulted in non-convergent results, so we adopted a small factor, $\damp=0.05$. For a fair comparison, we use the same factor for the visibility-based solutions. We then observe the error in the gain estimates introduced by the noise.
\begin{equation}\label{eq:gain_error}
\sigma_g = \left[\frac{1}{N_f \Nant} \sum_f \sum_a \frac{\left|g'_a(f)-g^T_a(f)\right|^2}{\left|g^T_a(f)\right|^2}\right]^{1/2}
\end{equation}

@@

\textbf{Improving sky model.} In this work we used a perfect sky model (for simulation), or a very simple sky model (for LWA data). In principle the images produced by the EPIC correlator can be used to improve the sky model used in calibration -- similar to a major loop in self-calibration. This can be especially useful for compact, widefield arrays which require a model of both compact and diffuse sources over a large patch of sky. In addition to requiring the sky model, the calibration also requires this model to be attenuated by the primary beam, which can be difficult to measure at the precision necessary \citep[e.g.][]{neb15,vir14,thy15b}. However, the direct imaging correlator provides exactly the model necessary in real time, and can be iterated over to improve the images and gain solutions.

\textbf{Dynamic parameters.} In our controlled experiments we fine tuned a number of parameters based on our testing (e.g. damping factor, integration time, frequency averaging). A deployed system will require robust determination of these parameters to operate continuously. The specifics will be heavily dependent on the stability of the instrument, the frequency of observation, and the sky. For example, a stable instrument may be able to use short integrations to determine a rough estimate of the gains before switching to much longer integration (on order seconds to minutes) to highly increase signal to noise. At low frequencies or high imaging resolution, the dynamics of the ionosphere are important, and will likely drive the limit of time integration allowed.

\textbf{Multiple pixel correlations.} As pointed out in section~\ref{sec:noise}, for a sky-noise dominated system, the signal to noise in the calibrator pixel dominates the error on the EPICal generated gain estimates. Besides the methods described above to increase signal to noise, one could also use correlations of multiple image pixels to incorporate a higher fraction of the total sky noise. Of course this increases the computational cost to a scaling of $\mathcal{O}(\Nant N_{\mathrm{pix}})$, which will typically be much lower than the $\mathcal{O}(\Ng \log_2 \Ng)$ of the correlator itself.

\textbf{Direction dependent gains.} Even below the timescale of a dynamic ionosphere, a widefield instrument may require direction dependent gains to account for a varying thickness ionospheric sheet. The holographic mapping framework used by the MOFF algorithm, and in turn the EPIC and EPICal software can naturally account for direction dependence through an updated model of the antenna voltage pattern. This can be achieved by using multiple pixel correlations to independently solve for gains in each pixel direction. These solutions can then be used to fit a beam model on the sky (analogous to the gain model fitting mentioned above), and updating the antenna voltage pattern. This updated pattern would then feed into the gridding step of the correlator, allowing the imager to convolve the signals with the effective beam pattern in the ground plane.
